---
title: "RTMB EBS pollock: standalone spmR projection attachment"
subtitle: "Projection inputs, seven Tier-3 alternatives, and reproducibility lineage"
author: "James Ianelli"
date: today
date-format: "MMMM D, YYYY"
bibliography: ../references/software.bib
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
    theme:
      light: cosmo
      dark: darkly
execute:
  echo: false
  warning: false
  message: false
---

{{< include _ai_disclosure.qmd >}}

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggthemes)
  library(gt)
})
theme_set(ggthemes::theme_few())

repo_root <- normalizePath("..", mustWork = TRUE)
projection_dir <- file.path(repo_root, "analysis", "output", "spmR_projection")
required_files <- file.path(
  projection_dir,
  c("base_lineage.csv", "projection_run_metadata.csv",
    "tier3_seven_scenario_table.csv", "spm_detail.csv")
)
if (any(!file.exists(required_files))) {
  stop("Run R/run_spmr_tier3_projection.R before rendering this attachment.")
}
lineage <- read.csv(required_files[1], check.names = FALSE)
metadata <- read.csv(required_files[2], check.names = FALSE)
tier3 <- read.csv(required_files[3], check.names = FALSE)
detail <- read.csv(required_files[4], check.names = FALSE)
if (metadata$spmR_version[1] != as.character(packageVersion("spmR"))) {
  stop("Installed spmR version differs from the projection-run metadata.")
}
if (normalizePath(metadata$model_file[1], mustWork = TRUE) !=
      normalizePath(lineage$model_file[1], mustWork = TRUE) ||
    unname(tools::md5sum(metadata$model_file[1])) != lineage$model_md5[1]) {
  stop("The saved RTMB fit differs from the fit used for these projections.")
}
alt_names <- setNames(paste0("Alt ", tier3$Alt, ": ", tier3$Scenario), tier3$Alt)
detail <- detail |>
  mutate(scenario = factor(alt_names[as.character(Alt)], levels = unname(alt_names)))
```

# Scope and lineage

This standalone attachment applies `spmR` [@ianelli2026spmR] to the saved RTMB
EBS pollock bridge. It embeds its tables and figures in one HTML file and does
not depend on the SPoRC or Rceattle applications. The accepted ADMB assessment
remains the management reference.

```{r}
#| label: tbl-rtmb-spmr-lineage
#| tbl-cap: "RTMB and spmR identifiers used for this projection attachment."
data.frame(
  Item = c("Model", "RTMB fit", "Fit MD5", "spmR version", "Alternatives",
           "Simulations", "Comparison years", "Generated"),
  Value = c("RTMB EBS pollock", metadata$model_file[1], lineage$model_md5[1],
            metadata$spmR_version[1], metadata$alternatives[1],
            metadata$simulations[1], metadata$comparison_years[1],
            metadata$generated[1])
) |>
  gt() |>
  tab_header(title = "RTMB Projection Lineage")
```

# Seven Tier-3 scenario results

```{r}
#| label: tbl-rtmb-spmr-seven-scenarios
#| tbl-cap: "Tier-3 projection results generated specifically from the RTMB EBS pollock bridge. Catch and biomass quantities are thousand t."
tier3 |>
  gt() |>
  fmt_number(columns = matches("^(Catch|ABC|OFL|SSB)_"), decimals = 0) |>
  fmt_number(columns = matches("^F_"), decimals = 3) |>
  fmt_number(columns = matches("^B_B35_"), decimals = 2) |>
  cols_label(
    Alt = "Alternative",
    Catch_2027 = "Catch", ABC_2027 = "ABC", OFL_2027 = "OFL",
    SSB_2027 = md("$B$"), F_2027 = md("$F$"), B_B35_2027 = md("$B/B_{35\\%}$"),
    Catch_2028 = "Catch", ABC_2028 = "ABC", OFL_2028 = "OFL",
    SSB_2028 = md("$B$"), F_2028 = md("$F$"), B_B35_2028 = md("$B/B_{35\\%}$")
  ) |>
  tab_spanner(label = "2027", columns = ends_with("_2027")) |>
  tab_spanner(label = "2028", columns = ends_with("_2028")) |>
  tab_header(title = "Seven Tier-3 Projection Alternatives")
```

# Projection trajectories

```{r}
#| label: fig-rtmb-spmr-biomass
#| fig-cap: "Projected spawning biomass by Tier-3 alternative. Lines show simulation medians and bands span the 5th to 95th percentiles; dashed lines mark B35%."
#| fig-alt: "Seven-panel projection of spawning biomass, with a median line, uncertainty band, and dashed B35-percent reference in each panel."
detail |>
  group_by(scenario, Year) |>
  summarize(
    lower = quantile(SSB, 0.05), median = median(SSB), upper = quantile(SSB, 0.95),
    B35 = first(B35), .groups = "drop"
  ) |>
  ggplot(aes(Year, median)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#2b7a9b", alpha = 0.22) +
  geom_line(color = "#102f3d", linewidth = 0.8) +
  geom_line(aes(y = B35, linetype = "B35%"), color = "#4d4d4d") +
  facet_wrap(~scenario, ncol = 2) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_linetype_manual(values = c("B35%" = "dashed")) +
  labs(x = "Year", y = "Spawning biomass (thousand t)", linetype = "Reference") +
  ggthemes::theme_few() +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-rtmb-spmr-removals
#| fig-cap: "Median projected catch, ABC, and OFL by Tier-3 alternative."
#| fig-alt: "Seven-panel projection chart with separate line types for median catch, ABC, and OFL."
detail |>
  select(scenario, Year, Catch, ABC, OFL) |>
  pivot_longer(c(Catch, ABC, OFL), names_to = "quantity", values_to = "value") |>
  group_by(scenario, Year, quantity) |>
  summarize(value = median(value), .groups = "drop") |>
  ggplot(aes(Year, value, linetype = quantity)) +
  geom_line(color = "#102f3d", linewidth = 0.8) +
  facet_wrap(~scenario, ncol = 2) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x = "Year", y = "Thousand metric tons", linetype = "Quantity") +
  ggthemes::theme_few() +
  theme(legend.position = "bottom")
```

# Reproduction

From the RTMB repository root:

```text
Rscript R/run_spmr_tier3_projection.R
quarto render reporting/appendix_spmr_projections.qmd
```
