---
title: "Fishery Selectivity Form Comparison"
execute:
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    lightbox: true
    # KaTeX is shipped with Quarto and works well with embed-resources (no external CDN reliance).
    html-math-method: katex
    code-fold: true
    code-summary: "Show code"
    code-tools: true
params:
  results_dir: "analysis/output/fishery_sel_forms"
  base_form: 0
  compare_forms: [2, 5]
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggthemes)
  library(ggridges)
  library(gt)
  library(knitr)
  library(tibble)
  library(purrr)
  library(readr)
})

theme_set(ggthemes::theme_few())

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0) y else x

# Render can run with working dir = reporting/, so climb one level if needed.
wd <- normalizePath(getwd(), mustWork = TRUE)
rtmb_root <- if (file.exists(file.path(wd, "R", "Rpm.R"))) wd else normalizePath(file.path(wd, ".."), mustWork = TRUE)
rtmb_file <- function(...) normalizePath(file.path(rtmb_root, ...), mustWork = FALSE)

results_dir <- rtmb_file(params$results_dir)

run_path <- function(form) file.path(results_dir, sprintf("fishery_sel_form_%d.rds", as.integer(form)))

load_run <- function(form) {
  path <- run_path(form)
  if (!file.exists(path)) stop("Missing results file: ", path)
  x <- readRDS(path)
  x$path <- path
  x
}

base_form <- as.integer(params$base_form)
compare_forms <- as.integer(params$compare_forms)
forms <- c(base_form, compare_forms)
runs <- purrr::map(forms, load_run)

safe_logit_rho <- function(x) 2 / (1 + exp(-x)) - 1

# Short labels for plotting (full labels can get clipped in facets)
short_label <- function(form) {
  form <- as.integer(form)
  if (form == 0L) return("Base")
  if (form == 2L) return("Double-logistic")
  if (form == 5L) return("2D AR1")
  paste0("Form ", form)
}

summary_tbl <- tibble(
  fishery_sel_form = purrr::map_int(runs, "form"),
  label = purrr::map_chr(runs, "label"),
  label_short = purrr::map_chr(purrr::map_int(runs, "form"), short_label),
  objective = purrr::map_dbl(runs, "objective"),
  evaluated_total = purrr::map_dbl(runs, ~ .x$evaluated_total %||% .x$report$tot_like %||% NA_real_),
  convergence = purrr::map_int(runs, "convergence"),
  max_gradient = purrr::map_dbl(runs, "max_gradient"),
  seconds = purrr::map_dbl(runs, "seconds")
) |>
  arrange(fishery_sel_form) |>
  mutate(
    delta_objective = objective - objective[fishery_sel_form == base_form][1],
    delta_evaluated_total = evaluated_total - evaluated_total[fishery_sel_form == base_form][1],
    objective_report_gap = evaluated_total - objective
  )

rho_ar1_summary <- {
  ar1_run <- runs[[which(purrr::map_int(runs, "form") == 5L)[1]]]
  rho_working <- ar1_run$fixed_parameters$sel_tv_ar1_rho_fsh %||% c(NA_real_, NA_real_)
  rho <- safe_logit_rho(as.numeric(rho_working))
  tibble(
    fishery_sel_form = 5L,
    label_short = short_label(5L),
    rho_y = rho[1],
    rho_a = rho[2],
    working_rho_y = as.numeric(rho_working[1]),
    working_rho_a = as.numeric(rho_working[2])
  )
}

rho_ar1_caption <- if (nrow(rho_ar1_summary) > 0 && all(is.finite(rho_ar1_summary$rho_y), is.finite(rho_ar1_summary$rho_a))) {
  sprintf(
    " For the 2D AR1 form, fitted correlations are rho_y = %.3f and rho_a = %.3f.",
    rho_ar1_summary$rho_y,
    rho_ar1_summary$rho_a
  )
} else {
  ""
}

get_ssb_df <- function(run) {
  ssb <- run$report$SSB
  tibble(
    fishery_sel_form = run$form,
    label = run$label,
    label_short = short_label(run$form),
    year = as.integer(names(ssb)),
    SSB = as.numeric(ssb)
  )
}

# Note: fishery selectivity is stored as sel_fsh in the report list (matrix [year x age])
get_sel_df <- function(run) {
  sel <- run$report$sel_fsh
  stopifnot(is.matrix(sel))
  years <- as.integer(rownames(sel) %||% seq_len(nrow(sel)))
  if (is.null(rownames(sel))) years <- seq_len(nrow(sel))
  out <- as_tibble(sel, .name_repair = "minimal")
  names(out) <- seq_len(ncol(sel))
  out |>
    mutate(year = years, .before = 1) |>
    pivot_longer(-year, names_to = "age", values_to = "selectivity") |>
    mutate(
      age = as.integer(age),
      fishery_sel_form = run$form,
      label = run$label,
      label_short = short_label(run$form)
    )
}

ssb_df <- bind_rows(purrr::map(runs, get_ssb_df))
sel_df <- bind_rows(purrr::map(runs, get_sel_df))

get_catch_df <- function(run) {
  obs <- run$report$obs_catch
  pred <- run$report$pred_catch
  yrs <- as.integer(names(obs) %||% names(pred) %||% seq_along(obs))
  tibble(
    fishery_sel_form = run$form,
    label = run$label,
    label_short = short_label(run$form),
    year = yrs,
    obs_catch = as.numeric(obs),
    pred_catch = as.numeric(pred)
  )
}

catch_df <- bind_rows(purrr::map(runs, get_catch_df))

# Generic GT wrapper

gt_report <- function(data, ...) {
  gt(data, ...) |>
    tab_options(table.font.size = px(13))
}

plot_agecomp_fit_comparison <- function(data, fleet_name, ncol = 4) {
  plot_data <- data |>
    filter(fleet == fleet_name, is.finite(obs) | is.finite(exp)) |>
    mutate(label_short = factor(label_short, levels = unique(summary_tbl$label_short)))

  observed_data <- plot_data |>
    group_by(fleet, year, age) |>
    summarize(obs = mean(obs, na.rm = TRUE), .groups = "drop")

  ggplot() +
    geom_line(
      data = plot_data,
      aes(x = age, y = exp, color = label_short, group = label_short),
      linewidth = 0.5,
      alpha = 0.9,
      na.rm = TRUE
    ) +
    geom_point(
      data = observed_data,
      aes(x = age, y = obs),
      color = "#1f78b4",
      size = 0.85,
      alpha = 0.75,
      na.rm = TRUE
    ) +
    facet_wrap(~year, ncol = ncol, dir = "v") +
    scale_y_continuous(limits = c(0, NA)) +
    labs(
      x = "Age",
      y = "Proportion",
      color = "Scenario",
      title = paste(fleet_name, "Age Composition Fits")
    ) +
    theme(legend.position = "top")
}
```

# Overview

This report compares three RTMB runs that differ **only** in the fishery selectivity form.

- Base: `fishery_sel_form = {params.base_form}`
- Comparators: `fishery_sel_form = {paste(params.compare_forms, collapse=', ')}`

The run outputs were generated by `R/run_fishery_selectivity_forms.R` and saved under:

- `{params.results_dir}`

# Fit Summary

```{r}
#| label: tbl-fit-summary
#| tbl-cap: "Model optimization summary by fishery selectivity form."
summary_tbl |>
  mutate(across(c(objective, evaluated_total, delta_objective, delta_evaluated_total,
                  objective_report_gap, max_gradient, seconds), ~round(.x, 6))) |>
  gt_report() |>
  tab_header(title = "Fit summary")
```

`objective_report_gap` should be near zero. A large value indicates that the optimizer
and the saved report were evaluated under different data or parameter settings.

```{r}
#| label: objective-report-gap-warning
#| echo: false
#| results: asis

gap_max <- max(abs(summary_tbl$objective_report_gap), na.rm = TRUE)
if (is.finite(gap_max) && gap_max > 1e-3) {
  cat(
    "\n\n> **Diagnostic warning:** at least one saved run has a large ",
    "`objective_report_gap`. Regenerate the scenario `.rds` files with the ",
    "current `R/run_fishery_selectivity_forms.R` before interpreting the ",
    "selectivity comparison.\n\n",
    sep = ""
  )
}
```

## Catch biomass fit (observed vs predicted)

```{r}
#| label: fig-catch-fit
#| fig-cap: "Observed vs predicted catch biomass by scenario (fishery)."

catch_long <- catch_df |>
  pivot_longer(c(obs_catch, pred_catch), names_to = "series", values_to = "catch") |>
  mutate(series = factor(series, levels = c("obs_catch", "pred_catch"), labels = c("Observed", "Predicted")))

ggplot(catch_long, aes(x = year, y = catch, color = series)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~label_short, ncol = 1, scales = "free_y") +
  labs(x = "Year", y = "Catch biomass", color = NULL) +
  theme(legend.position = "bottom")
```

# Likelihood Component Comparison

## Age-composition fit comparison

The model returns an age-composition negative log-likelihood contribution by fleet in `report$age_like`.

It also stores fishery observed (`oac_fsh`) and expected (`eac_fsh`) age-composition proportions by year×age in the saved run outputs.

```{r}
#| label: fig-age-like
#| fig-cap: "Change in age-composition NLL relative to the base scenario by fleet. Negative values indicate improved fit relative to base."
#| fig-width: 8
#| fig-height: 4.5
#| out-width: "100%"

age_like_df <- bind_rows(purrr::map(runs, function(run) {
  al <- run$report$age_like
  tibble(
    fishery_sel_form = run$form,
    label = run$label,
    label_short = short_label(run$form),
    fleet = names(al),
    age_like = as.numeric(al)
  )
})) |>
  mutate(
    fleet = factor(fleet, levels = c("fsh", "bts", "ats"), labels = c("Fishery", "BTS", "ATS")),
    label = factor(label, levels = unique(summary_tbl$label)),
    label_short = factor(label_short, levels = unique(summary_tbl$label_short))
  )

base_vals <- age_like_df |>
  filter(fishery_sel_form == base_form) |>
  select(fleet, base_age_like = age_like)

age_like_df2 <- age_like_df |>
  left_join(base_vals, by = "fleet") |>
  mutate(delta_age_like = age_like - base_age_like)

age_like_plot_df <- age_like_df2 |>
  filter(fishery_sel_form != base_form)

ggplot(age_like_plot_df, aes(x = label_short, y = delta_age_like, fill = label_short)) +
  geom_hline(yintercept = 0, linewidth = 0.4, color = "gray50") +
  geom_col(width = 0.68, alpha = 0.9) +
  geom_text(
    aes(
      label = round(delta_age_like, 1),
      vjust = ifelse(delta_age_like >= 0, -0.35, 1.25)
    ),
    size = 3,
    show.legend = FALSE
  ) +
  facet_wrap(~fleet, nrow = 1, scales = "free_y") +
  labs(x = NULL, y = "Delta age-comp NLL vs base", fill = NULL) +
  theme(legend.position = "top")
```

```{r}
#| label: tbl-age-like
#| tbl-cap: "Age-composition NLL contributions (and deltas vs base) by fleet and scenario."

age_like_df2 |>
  mutate(across(c(age_like, base_age_like, delta_age_like), ~round(.x, 0))) |>
  arrange(fleet, fishery_sel_form) |>
  gt_report()
```

## Age Composition Fits

These figures follow the observed-versus-expected age-composition layout used for figures 5-7 in the main assessment report. Points show the observed proportions; colored lines show the expected proportions from each fishery selectivity scenario.

```{r}
#| label: agecomp-data
#| include: false

extract_agecomp_df <- function(run, fleet = c("fsh", "bts", "ats")) {
  fleet <- match.arg(fleet)

  oac <- run$report[[paste0("oac_", fleet)]]
  eac <- run$report[[paste0("phat_", fleet)]]
  stopifnot(!is.null(oac), !is.null(eac))

  yrs_key <- paste0("yrs_", fleet, "_data")
  yrs <- run$report[[yrs_key]]
  if (is.null(yrs) || length(yrs) != nrow(oac)) {
    # Fallback if years weren't saved (shouldn't happen): use row index
    yrs <- seq_len(nrow(oac))
  }

  sample_size <- run$report[[paste0("sam_", fleet)]]
  if (is.null(sample_size) || length(sample_size) != nrow(oac)) {
    sample_size <- rep(NA_real_, nrow(oac))
  }

  ages <- seq_len(ncol(oac))

  # Match the main report's age-composition plots by comparing BTS/ATS over ages 2+.
  if (fleet %in% c("bts", "ats") && ncol(oac) >= 2) {
    keep <- 2:ncol(oac)
    oac <- oac[, keep, drop = FALSE]
    eac <- eac[, keep, drop = FALSE]
    ages <- keep
  }

  grid <- tidyr::expand_grid(year = yrs, age = ages) |>
    as.data.frame()
  grid$fishery_sel_form <- run$form
  grid$label <- run$label
  grid$label_short <- short_label(run$form)
  grid$fleet <- fleet

  # Convert to proportions within each year-row to make diagnostics comparable.
  # Some ATS expected values are stored on an abundance scale before this normalization.
  o_rs <- pmax(rowSums(oac), 1e-12)
  e_rs <- pmax(rowSums(eac), 1e-12)
  o_prop <- sweep(oac, 1, o_rs, "/")
  e_prop <- sweep(eac, 1, e_rs, "/")

  # Safety: enforce exact renormalization numerically (and catch pathologies)
  o_prop <- sweep(o_prop, 1, pmax(rowSums(o_prop), 1e-12), "/")
  e_prop <- sweep(e_prop, 1, pmax(rowSums(e_prop), 1e-12), "/")

  # Optional checks (won't stop render, but will warn in console if something is off)
  o_chk <- max(abs(rowSums(o_prop) - 1))
  e_chk <- max(abs(rowSums(e_prop) - 1))
  if (is.finite(o_chk) && o_chk > 1e-6) warning("oac proportions do not sum to 1 (max abs dev = ", o_chk, ") for fleet=", fleet, ", form=", run$form)
  if (is.finite(e_chk) && e_chk > 1e-6) warning("eac proportions do not sum to 1 (max abs dev = ", e_chk, ") for fleet=", fleet, ", form=", run$form)

  grid$obs <- as.numeric(t(o_prop))
  grid$exp <- as.numeric(t(e_prop))
  grid$resid <- grid$obs - grid$exp
  grid$n_eff <- rep(as.numeric(sample_size), each = length(ages))

  tibble::as_tibble(grid)
}

agecomp_df <- bind_rows(
  purrr::map_dfr(runs, extract_agecomp_df, fleet = "fsh"),
  purrr::map_dfr(runs, extract_agecomp_df, fleet = "bts"),
  purrr::map_dfr(runs, extract_agecomp_df, fleet = "ats")
) |>
  mutate(
    label = factor(label, levels = unique(summary_tbl$label)),
    label_short = factor(label_short, levels = unique(summary_tbl$label_short)),
    fleet = factor(fleet, levels = c("fsh", "bts", "ats"), labels = c("Fishery", "BTS", "ATS"))
  )
```

```{r}
#| label: fig-agecomp-fishery
#| fig-cap: "Observed and expected fishery age-composition patterns by fishery selectivity scenario, faceted by year. Points are observed proportions; lines are scenario-specific expected proportions."
#| fig-width: 8
#| fig-height: 11
#| out-width: "100%"
plot_agecomp_fit_comparison(agecomp_df, "Fishery", ncol = 5)
```

```{r}
#| label: fig-agecomp-bts
#| fig-cap: "Observed and expected BTS age-composition patterns by fishery selectivity scenario, faceted by year. Points are observed proportions; lines are scenario-specific expected proportions."
#| fig-width: 8
#| fig-height: 11
#| out-width: "100%"
plot_agecomp_fit_comparison(agecomp_df, "BTS", ncol = 4)
```

```{r}
#| label: fig-agecomp-ats
#| fig-cap: "Observed and expected ATS age-composition patterns by fishery selectivity scenario, faceted by year. Points are observed proportions; lines are scenario-specific expected proportions."
#| fig-width: 8
#| fig-height: 11
#| out-width: "100%"
plot_agecomp_fit_comparison(agecomp_df, "ATS", ncol = 3)
```

The RTMB report object stores a set of likelihood (or penalty) components. The most relevant ones for this comparison are:

- `sel_like`, `sel_like_dev` (selectivity penalties)
- `age_like` (age composition likelihood)
- `cat_like` (catch fit)
- `cpue_like` (CPUE)

```{r}
#| label: tbl-nll-components
#| tbl-cap: "Selected likelihood / penalty components returned in the RTMB report."

extract_components <- function(run) {
  r <- run$report
  tibble(
    fishery_sel_form = run$form,
    label = run$label,
    sel_like = sum(r$sel_like, na.rm = TRUE),
    sel_like_dev = sum(r$sel_like_dev, na.rm = TRUE),
    age_like_total = attr(r$age_like, "total") %||% sum(r$age_like, na.rm = TRUE),
    cat_like = r$cat_like %||% NA_real_,
    cpue_like = r$cpue_like %||% NA_real_,
    avo_like = r$avo_like %||% NA_real_,
    wt_like = r$wt_like %||% NA_real_,
    nll = r$tot_like %||% r$nll %||% NA_real_
  )
}

comp_tbl <- bind_rows(purrr::map(runs, extract_components)) |>
  arrange(fishery_sel_form) |>
  mutate(
    delta_nll = nll - nll[fishery_sel_form == base_form][1],
    delta_sel_like = sel_like - sel_like[fishery_sel_form == base_form][1],
    delta_sel_like_dev = sel_like_dev - sel_like_dev[fishery_sel_form == base_form][1]
  )

comp_tbl |>
  mutate(across(where(is.numeric), ~round(.x, 0))) |>
  gt_report() |>
  tab_header(title = "NLL components")
```

# Fishery Selectivity Surfaces

Fishery selectivity is represented as a relative age pattern that multiplies annual fully selected fishing mortality. Let \(t\) index years, \(a\) index ages, and \(x_a = a + 0.5\) denote the age value used in the selectivity functions. Each form is converted to log selectivity \(\ell_{t,a}\), constrained so ages 11 and older share the age-11 value, centered within year, and exponentiated:

$$
\ell_{t,a} = \ell_{t,11}, \quad a \ge 11,
$$

$$
\ell^*_{t,a} =
\ell_{t,a} -
\log \left\{ A^{-1} \sum_{a=1}^{A} \exp(\ell_{t,a}) \right\},
\qquad
s_{t,a} = \exp(\ell^*_{t,a}).
$$

The centered selectivity \(s_{t,a}\) enters the population dynamics as \(F_{t,a} = F_t s_{t,a}\). Code references needed to reproduce the implementation are in @sec-appendix-implementation.

## Form 0: base coefficients with change-year deviations

The base model uses estimated age coefficients plus deviations at configured fishery-selectivity change years:

$$
\ell_{t,a} = c_a + d_{k(t),a},
$$

where \(c_a\) is the baseline age coefficient and \(d_{k(t),a}\) is the deviation for the change-year block \(k(t)\). This form is flexible at the configured break points but does not estimate a continuous year-to-year process for fishery selectivity.

## Form 2: time-varying double logistic

The double-logistic form estimates annual parameters \(p_{1,t}\), \(p_{2,t}\), and \(p_{3,t}\), with starting values centered on \(1.5\), \(3\), and \(4\). These parameters define the ascending and descending transition ages:

$$
\gamma_{1,t} = p_{1,t} + p_{2,t},
\qquad
\gamma_{2,t} = 2p_{1,t} + p_{2,t} + p_{3,t}.
$$

The uncentered annual curve is

$$
u_{t,a} =
\left[1 + \exp\left\{-\log(19)\frac{x_a - \gamma_{1,t}}{p_{1,t}}\right\}\right]^{-1},
$$

$$
v_{t,a} =
1 -
\left[1 + \exp\left\{-\log(19)\frac{x_a - \gamma_{2,t}}{p_{3,t}}\right\}\right]^{-1},
$$

$$
\ell_{t,a} =
\log\left(u_{t,a} v_{t,a} 0.95^{-2} + \epsilon \right)
- \max_a \log\left(u_{t,a} v_{t,a} 0.95^{-2} + \epsilon \right),
$$

followed by the age-11 plus-group constraint and within-year centering. The annual log-scale triplets \(\theta_t = \log(p_{1,t}, p_{2,t}, p_{3,t})\) are penalized with shrinkage toward \(\mu = \log(1.5, 3, 4)\) and a first-difference penalty:

$$
Q_\mathrm{DL} =
\sum_t \sum_j
\left(\frac{\theta_{t,j} - \mu_j}{\sigma_{p,j}}\right)^2
+
\frac{1}{2}
\sum_{t=2}^{T} \sum_j
\left(\frac{\theta_{t,j} - \theta_{t-1,j}}{\sigma_{\Delta,j}}\right)^2,
$$

where \(\sigma_p = (0.75, 0.75, 0.45)\) and \(\sigma_\Delta = (0.60, 0.60, 0.40)\). The smaller scales for \(p_3\) impose stronger regularization on the descending limb, which is otherwise weakly informed by older ages.

## Form 5: 2D AR1 time-varying surface

The 2D AR1 form estimates a full year-by-age latent field \(z_{t,a}\). Before centering, the log selectivity is

$$
\ell_{t,a} = z_{t,a},
$$

with the same age-11 plus-group constraint and within-year centering used for the other forms. The latent field is penalized with a separable AR1 process over year and age:

$$
z \sim \operatorname{AR1}_{\mathrm{year}}(\rho_y)
\otimes
\operatorname{AR1}_{\mathrm{age}}(\rho_a),
$$

with fitted correlations reported in the caption for @fig-sel-ridges. The scenario recipe down-weights the AR1 penalty relative to the default, allowing more interannual variation while still borrowing strength across neighboring years and ages.

## Selectivity curves (selected years)

```{r}
#| label: fig-sel-curves
#| fig-cap: "Fishery selectivity-at-age curves for selected years, by fishery selectivity form."

years_to_show <- sort(unique(sel_df$year))
years_to_show <- years_to_show[years_to_show %in% c(min(years_to_show), 1980, 1995, 2010, 2024)]
if (length(years_to_show) == 0) years_to_show <- head(sort(unique(sel_df$year)), 5)

plot_df <- sel_df |>
  filter(year %in% years_to_show) |>
  group_by(fishery_sel_form, label, year) |>
  mutate(relative = selectivity / max(selectivity, na.rm = TRUE)) |>
  ungroup()

ggplot(plot_df, aes(x = age, y = relative, color = factor(year))) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~label_short, ncol = 1, scales = "fixed") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = "Age", y = "Relative selectivity", color = "Year")
```

## ggridges (all years)

```{r}
#| label: fig-sel-ridges
#| fig-cap: !expr paste0("Fishery selectivity through time (ggridges), normalized within year.", rho_ar1_caption)
#| fig-width: 9
#| fig-height: 11
#| out-width: "100%"

ridge_df <- sel_df |>
  group_by(fishery_sel_form, label, year) |>
  mutate(relative = selectivity / max(selectivity, na.rm = TRUE)) |>
  ungroup() |>
  mutate(year_factor = factor(year, levels = rev(sort(unique(year)))))

ggplot(ridge_df, aes(x = age, y = year_factor, height = relative, group = year_factor)) +
  geom_density_ridges(stat = "identity", fill = "#1f78b4", alpha = 0.65, rel_min_height = 0.01) +
  facet_wrap(~label_short, ncol = 3) +
  labs(x = "Age", y = "Year") +
  theme(axis.text.y = element_text(size = 7))
```

# Spawning Biomass Comparison

```{r}
#| label: fig-ssb
#| fig-cap: "Spawning biomass (SSB) trajectories by fishery selectivity form."

ggplot(ssb_df, aes(x = year, y = SSB, color = label)) +
  geom_line(linewidth = 1) +
  labs(x = "Year", y = "SSB", color = NULL) +
  theme(legend.position = "top")
```

```{r}
#| label: tbl-ssb-differences
#| tbl-cap: "SSB differences relative to base run at key years."

key_years <- sort(unique(ssb_df$year))
key_years <- key_years[key_years %in% c(max(key_years), max(key_years) - 1, 2000, 2010, 2020)]
if (length(key_years) == 0) key_years <- tail(sort(unique(ssb_df$year)), 5)

ssb_wide <- ssb_df |>
  filter(year %in% key_years) |>
  select(fishery_sel_form, label, year, SSB) |>
  pivot_wider(names_from = fishery_sel_form, values_from = SSB, names_prefix = "form_")

base_col <- paste0("form_", base_form)

ssb_delta <- ssb_df |>
  filter(year %in% key_years) |>
  group_by(year) |>
  mutate(
    base_SSB = SSB[fishery_sel_form == base_form][1],
    delta_SSB = SSB - base_SSB,
    pct_delta = 100 * delta_SSB / base_SSB
  ) |>
  ungroup() |>
  select(label, fishery_sel_form, year, SSB, delta_SSB, pct_delta) |>
  arrange(year, fishery_sel_form)

ssb_delta |>
  mutate(across(c(SSB, delta_SSB, pct_delta), ~round(.x, 4))) |>
  gt_report() |>
  tab_header(title = "SSB deltas vs base")
```

# Notes

- The age-composition panels above now use the same observed-point / expected-line structure as the main assessment report, with expected lines overlaid by fishery selectivity scenario.

# Appendix: Implementation References {#sec-appendix-implementation}

The equations above are implemented through the RTMB parameter builder, the selectivity construction helper, and the model likelihood. These references provide traceability without repeating the methods section.

## Parameter blocks

Fishery selectivity parameters are initialized in `R/utils-rtmb.R::add_fishery_selectivity_parameters()`. Defaults include the annual double-logistic triplets for form 2 (`sel_double_logistic_fsh`), the year-by-age latent field for form 5 (`sel_tv_ar1_fsh`), the two AR1 correlation parameters (`sel_tv_ar1_rho_fsh`), and the AR1 process scale (`log_sel_tv_ar1_sigma_fsh`). `R/Rpm.R` includes matching safety defaults so older saved runs can still be evaluated.

## Selectivity construction

`R/model_funs.R::compute_selectivity_fsh_forms()` dispatches among the fishery selectivity forms and returns the year-by-age log selectivity matrix. For form 0, it combines age coefficients and change-year deviations. For form 2, it evaluates the annual double-logistic curve. For form 5, it uses the full `sel_tv_ar1_fsh` latent field directly. The helper also applies the old-age plus-group constraint before the log selectivity is centered and returned.

`R/Rpm.R` then exponentiates the returned log selectivity and applies it inside the annual fishing mortality calculation:

```r
sel_fsh <- exp(log_sel_fsh)
F[i, ] <- Fmort[i] * sel_fsh[i, ]
```

Thus, all scenario differences shown in the comparison report enter through the same population-dynamics pathway after the selectivity surface has been constructed.

## Selectivity penalties

`R/model_funs.R::selectivity_like_fsh_forms()` evaluates the selectivity penalties. For form 2, it applies the annual log-parameter prior and first-difference penalty on \(p_1\), \(p_2\), and \(p_3\). For form 5, it transforms the estimated AR1 correlation parameters, evaluates the separable AR1-by-AR1 density for the year-by-age field, and applies the configured penalty weight.

## Scenario driver

`R/run_fishery_selectivity_forms.R::run_fishery_selectivity_forms()` defines the set of fishery selectivity scenarios and maps or frees the corresponding parameters before optimization. The report reads the saved outputs from `analysis/output/fishery_sel_forms/`, so the figures and tables reflect the fitted scenario objects rather than recomputing the model inside the document.
