---
title: "Fishery Selectivity Form Comparison"
execute:
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    lightbox: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
params:
  results_dir: "analysis/output/fishery_sel_forms"
  base_form: 0
  compare_forms: [2, 5]
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggthemes)
  library(ggridges)
  library(gt)
  library(knitr)
  library(tibble)
  library(purrr)
  library(readr)
})

theme_set(ggthemes::theme_few())

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0) y else x

# Render can run with working dir = reporting/, so climb one level if needed.
wd <- normalizePath(getwd(), mustWork = TRUE)
rtmb_root <- if (file.exists(file.path(wd, "R", "Rpm.R"))) wd else normalizePath(file.path(wd, ".."), mustWork = TRUE)
rtmb_file <- function(...) normalizePath(file.path(rtmb_root, ...), mustWork = FALSE)

results_dir <- rtmb_file(params$results_dir)

run_path <- function(form) file.path(results_dir, sprintf("fishery_sel_form_%d.rds", as.integer(form)))

load_run <- function(form) {
  path <- run_path(form)
  if (!file.exists(path)) stop("Missing results file: ", path)
  x <- readRDS(path)
  x$path <- path
  x
}

base_form <- as.integer(params$base_form)
compare_forms <- as.integer(params$compare_forms)
forms <- c(base_form, compare_forms)
runs <- purrr::map(forms, load_run)

summary_tbl <- tibble(
  fishery_sel_form = purrr::map_int(runs, "form"),
  label = purrr::map_chr(runs, "label"),
  objective = purrr::map_dbl(runs, "objective"),
  convergence = purrr::map_int(runs, "convergence"),
  max_gradient = purrr::map_dbl(runs, "max_gradient"),
  seconds = purrr::map_dbl(runs, "seconds")
) |>
  arrange(fishery_sel_form) |>
  mutate(delta_objective = objective - objective[fishery_sel_form == base_form][1])

get_ssb_df <- function(run) {
  ssb <- run$report$SSB
  tibble(
    fishery_sel_form = run$form,
    label = run$label,
    year = as.integer(names(ssb)),
    SSB = as.numeric(ssb)
  )
}

# Note: fishery selectivity is stored as sel_fsh in the report list (matrix [year x age])
get_sel_df <- function(run) {
  sel <- run$report$sel_fsh
  stopifnot(is.matrix(sel))
  years <- as.integer(rownames(sel) %||% seq_len(nrow(sel)))
  if (is.null(rownames(sel))) years <- seq_len(nrow(sel))
  out <- as_tibble(sel, .name_repair = "minimal")
  names(out) <- seq_len(ncol(sel))
  out |>
    mutate(year = years, .before = 1) |>
    pivot_longer(-year, names_to = "age", values_to = "selectivity") |>
    mutate(
      age = as.integer(age),
      fishery_sel_form = run$form,
      label = run$label
    )
}

ssb_df <- bind_rows(purrr::map(runs, get_ssb_df))
sel_df <- bind_rows(purrr::map(runs, get_sel_df))

# Generic GT wrapper

gt_report <- function(data, ...) {
  gt(data, ...) |>
    tab_options(table.font.size = px(13))
}
```

# Overview

This report compares three RTMB runs that differ **only** in the fishery selectivity form.

- Base: `fishery_sel_form = {params.base_form}`
- Comparators: `fishery_sel_form = {paste(params.compare_forms, collapse=', ')}`

The run outputs were generated by `R/run_fishery_selectivity_forms.R` and saved under:

- `{params.results_dir}`

# Fit Summary

```{r}
#| label: tbl-fit-summary
#| tbl-cap: "Model optimization summary by fishery selectivity form."
summary_tbl |>
  mutate(across(c(objective, delta_objective, max_gradient, seconds), ~round(.x, 6))) |>
  gt_report() |>
  tab_header(title = "Fit summary")
```

# Likelihood Component Comparison

The RTMB report object stores a set of likelihood (or penalty) components. The most relevant ones for this comparison are:

- `sel_like`, `sel_like_dev` (selectivity penalties)
- `age_like` (age composition likelihood)
- `cat_like` (catch fit)
- `cpue_like` (CPUE)

```{r}
#| label: tbl-nll-components
#| tbl-cap: "Selected likelihood / penalty components returned in the RTMB report."

extract_components <- function(run) {
  r <- run$report
  tibble(
    fishery_sel_form = run$form,
    label = run$label,
    sel_like = sum(r$sel_like, na.rm = TRUE),
    sel_like_dev = sum(r$sel_like_dev, na.rm = TRUE),
    age_like_total = attr(r$age_like, "total") %||% sum(r$age_like, na.rm = TRUE),
    cat_like = r$cat_like %||% NA_real_,
    cpue_like = r$cpue_like %||% NA_real_,
    avo_like = r$avo_like %||% NA_real_,
    wt_like = r$wt_like %||% NA_real_,
    nll = r$nll %||% NA_real_
  )
}

comp_tbl <- bind_rows(purrr::map(runs, extract_components)) |>
  arrange(fishery_sel_form) |>
  mutate(
    delta_nll = nll - nll[fishery_sel_form == base_form][1],
    delta_sel_like = sel_like - sel_like[fishery_sel_form == base_form][1],
    delta_sel_like_dev = sel_like_dev - sel_like_dev[fishery_sel_form == base_form][1]
  )

comp_tbl |>
  mutate(across(where(is.numeric), ~round(.x, 6))) |>
  gt_report() |>
  tab_header(title = "NLL components")
```

# Fishery Selectivity Surfaces

## Selectivity curves (selected years)

```{r}
#| label: fig-sel-curves
#| fig-cap: "Fishery selectivity-at-age curves for selected years, by fishery selectivity form."

years_to_show <- sort(unique(sel_df$year))
years_to_show <- years_to_show[years_to_show %in% c(min(years_to_show), 1980, 1995, 2010, 2024)]
if (length(years_to_show) == 0) years_to_show <- head(sort(unique(sel_df$year)), 5)

plot_df <- sel_df |>
  filter(year %in% years_to_show) |>
  group_by(fishery_sel_form, label, year) |>
  mutate(relative = selectivity / max(selectivity, na.rm = TRUE)) |>
  ungroup()

ggplot(plot_df, aes(x = age, y = relative, color = factor(year))) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~label, ncol = 3, scales = "fixed") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = "Age", y = "Relative selectivity", color = "Year")
```

## ggridges (all years)

```{r}
#| label: fig-sel-ridges
#| fig-cap: "Fishery selectivity through time (ggridges), normalized within year."

ridge_df <- sel_df |>
  group_by(fishery_sel_form, label, year) |>
  mutate(relative = selectivity / max(selectivity, na.rm = TRUE)) |>
  ungroup() |>
  mutate(year_factor = factor(year, levels = rev(sort(unique(year)))))

ggplot(ridge_df, aes(x = age, y = year_factor, height = relative, group = year_factor)) +
  geom_density_ridges(stat = "identity", fill = "#1f78b4", alpha = 0.65, rel_min_height = 0.01) +
  facet_wrap(~label, ncol = 3) +
  labs(x = "Age", y = "Year") +
  theme(axis.text.y = element_text(size = 6))
```

# Spawning Biomass Comparison

```{r}
#| label: fig-ssb
#| fig-cap: "Spawning biomass (SSB) trajectories by fishery selectivity form."

ggplot(ssb_df, aes(x = year, y = SSB, color = label)) +
  geom_line(linewidth = 1) +
  labs(x = "Year", y = "SSB", color = NULL)
```

```{r}
#| label: tbl-ssb-differences
#| tbl-cap: "SSB differences relative to base run at key years."

key_years <- sort(unique(ssb_df$year))
key_years <- key_years[key_years %in% c(max(key_years), max(key_years) - 1, 2000, 2010, 2020)]
if (length(key_years) == 0) key_years <- tail(sort(unique(ssb_df$year)), 5)

ssb_wide <- ssb_df |>
  filter(year %in% key_years) |>
  select(fishery_sel_form, label, year, SSB) |>
  pivot_wider(names_from = fishery_sel_form, values_from = SSB, names_prefix = "form_")

base_col <- paste0("form_", base_form)

ssb_delta <- ssb_df |>
  filter(year %in% key_years) |>
  group_by(year) |>
  mutate(
    base_SSB = SSB[fishery_sel_form == base_form][1],
    delta_SSB = SSB - base_SSB,
    pct_delta = 100 * delta_SSB / base_SSB
  ) |>
  ungroup() |>
  select(label, fishery_sel_form, year, SSB, delta_SSB, pct_delta) |>
  arrange(year, fishery_sel_form)

ssb_delta |>
  mutate(across(c(SSB, delta_SSB, pct_delta), ~round(.x, 4))) |>
  gt_report() |>
  tab_header(title = "SSB deltas vs base")
```

# Notes

- If you want, we can add more diagnostics: fishery age-comp fits, residual bubbles, and/or likelihood-by-fleet breakdowns.
