---
title: "Fishery Selectivity Form Comparison"
execute:
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    lightbox: true
    html-math-method: katex
    code-fold: true
    code-summary: "Show code"
    code-tools: true
params:
  results_dir: "analysis/output/fishery_sel_forms"
  base_form: 0
  compare_forms: [2, 5]
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggthemes)
  library(ggridges)
  library(gt)
  library(knitr)
  library(tibble)
  library(purrr)
  library(readr)
})

theme_set(ggthemes::theme_few())

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0) y else x

# Render can run with working dir = reporting/, so climb one level if needed.
wd <- normalizePath(getwd(), mustWork = TRUE)
rtmb_root <- if (file.exists(file.path(wd, "R", "Rpm.R"))) wd else normalizePath(file.path(wd, ".."), mustWork = TRUE)
rtmb_file <- function(...) normalizePath(file.path(rtmb_root, ...), mustWork = FALSE)

results_dir <- rtmb_file(params$results_dir)

run_path <- function(form) file.path(results_dir, sprintf("fishery_sel_form_%d.rds", as.integer(form)))

load_run <- function(form) {
  path <- run_path(form)
  if (!file.exists(path)) stop("Missing results file: ", path)
  x <- readRDS(path)
  x$path <- path
  x
}

base_form <- as.integer(params$base_form)
compare_forms <- as.integer(params$compare_forms)
forms <- c(base_form, compare_forms)
runs <- purrr::map(forms, load_run)

summary_tbl <- tibble(
  fishery_sel_form = purrr::map_int(runs, "form"),
  label = purrr::map_chr(runs, "label"),
  objective = purrr::map_dbl(runs, "objective"),
  convergence = purrr::map_int(runs, "convergence"),
  max_gradient = purrr::map_dbl(runs, "max_gradient"),
  seconds = purrr::map_dbl(runs, "seconds")
) |>
  arrange(fishery_sel_form) |>
  mutate(delta_objective = objective - objective[fishery_sel_form == base_form][1])

get_ssb_df <- function(run) {
  ssb <- run$report$SSB
  tibble(
    fishery_sel_form = run$form,
    label = run$label,
    year = as.integer(names(ssb)),
    SSB = as.numeric(ssb)
  )
}

# Note: fishery selectivity is stored as sel_fsh in the report list (matrix [year x age])
get_sel_df <- function(run) {
  sel <- run$report$sel_fsh
  stopifnot(is.matrix(sel))
  years <- as.integer(rownames(sel) %||% seq_len(nrow(sel)))
  if (is.null(rownames(sel))) years <- seq_len(nrow(sel))
  out <- as_tibble(sel, .name_repair = "minimal")
  names(out) <- seq_len(ncol(sel))
  out |>
    mutate(year = years, .before = 1) |>
    pivot_longer(-year, names_to = "age", values_to = "selectivity") |>
    mutate(
      age = as.integer(age),
      fishery_sel_form = run$form,
      label = run$label
    )
}

ssb_df <- bind_rows(purrr::map(runs, get_ssb_df))
sel_df <- bind_rows(purrr::map(runs, get_sel_df))

# Generic GT wrapper

gt_report <- function(data, ...) {
  gt(data, ...) |>
    tab_options(table.font.size = px(13))
}
```

# Overview

This report compares three RTMB runs that differ **only** in the fishery selectivity form.

- Base: `fishery_sel_form = {params.base_form}`
- Comparators: `fishery_sel_form = {paste(params.compare_forms, collapse=', ')}`

The run outputs were generated by `R/run_fishery_selectivity_forms.R` and saved under:

- `{params.results_dir}`

# Fit Summary

```{r}
#| label: tbl-fit-summary
#| tbl-cap: "Model optimization summary by fishery selectivity form."
summary_tbl |>
  mutate(across(c(objective, delta_objective, max_gradient, seconds), ~round(.x, 6))) |>
  gt_report() |>
  tab_header(title = "Fit summary")
```

# Likelihood Component Comparison

## Age-composition fit comparison

The model returns an age-composition negative log-likelihood contribution by fleet in `report$age_like`.

It also stores fishery observed (`oac_fsh`) and expected (`eac_fsh`) age-composition proportions by year×age in the saved run outputs.

```{r}
#| label: fig-age-like
#| fig-cap: "Age-composition NLL contributions by fleet, across fishery selectivity scenarios."

age_like_df <- bind_rows(purrr::map(runs, function(run) {
  al <- run$report$age_like
  tibble(
    fishery_sel_form = run$form,
    label = run$label,
    fleet = names(al),
    age_like = as.numeric(al)
  )
})) |>
  mutate(
    fleet = factor(fleet, levels = c("fsh", "bts", "ats"), labels = c("Fishery", "BTS", "ATS")),
    label = factor(label, levels = unique(summary_tbl$label))
  )

base_vals <- age_like_df |>
  filter(fishery_sel_form == base_form) |>
  select(fleet, base_age_like = age_like)

age_like_df2 <- age_like_df |>
  left_join(base_vals, by = "fleet") |>
  mutate(delta_age_like = age_like - base_age_like)

p1 <- ggplot(age_like_df2, aes(x = fleet, y = age_like, fill = label)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  labs(x = NULL, y = "Age-comp NLL", fill = NULL) +
  theme(legend.position = "bottom")

p2 <- ggplot(age_like_df2, aes(x = fleet, y = delta_age_like, fill = label)) +
  geom_hline(yintercept = 0, linewidth = 0.4, color = "gray50") +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  labs(x = NULL, y = "Delta vs base", fill = NULL) +
  theme(legend.position = "bottom")

patchwork::wrap_plots(p1, p2, ncol = 1, guides = "collect")
```

```{r}
#| label: tbl-age-like
#| tbl-cap: "Age-composition NLL contributions (and deltas vs base) by fleet and scenario."

age_like_df2 |>
  mutate(across(c(age_like, base_age_like, delta_age_like), ~round(.x, 6))) |>
  arrange(fleet, fishery_sel_form) |>
  gt_report()
```

```{r}
#| label: fig-agecomp-bubbles
#| fig-cap: "Age-comp residual bubble plots (Observed − Expected) by fleet and scenario."

extract_agecomp_df <- function(run, fleet = c("fsh", "bts", "ats")) {
  fleet <- match.arg(fleet)

  oac <- run$report[[paste0("oac_", fleet)]]
  eac <- run$report[[paste0("phat_", fleet)]]
  stopifnot(!is.null(oac), !is.null(eac))

  yrs_key <- paste0("yrs_", fleet, "_data")
  yrs <- run$report[[yrs_key]]
  if (is.null(yrs) || length(yrs) != nrow(oac)) {
    # Fallback if years weren't saved (shouldn't happen): use row index
    yrs <- seq_len(nrow(oac))
  }

  ages <- seq_len(ncol(oac))

  grid <- tidyr::expand_grid(year = yrs, age = ages) |>
    as.data.frame()
  grid$fishery_sel_form <- run$form
  grid$label <- run$label
  grid$fleet <- fleet
  # Convert to proportions within each year-row to make residuals comparable
  o_prop <- sweep(oac, 1, pmax(rowSums(oac), 1e-12), "/")
  e_prop <- sweep(eac, 1, pmax(rowSums(eac), 1e-12), "/")

  grid$obs <- as.numeric(o_prop)
  grid$exp <- as.numeric(e_prop)
  grid$resid <- grid$obs - grid$exp
  tibble::as_tibble(grid)
}

agecomp_df <- bind_rows(
  purrr::map_dfr(runs, extract_agecomp_df, fleet = "fsh"),
  purrr::map_dfr(runs, extract_agecomp_df, fleet = "bts"),
  purrr::map_dfr(runs, extract_agecomp_df, fleet = "ats")
) |>
  mutate(
    label = factor(label, levels = unique(summary_tbl$label)),
    fleet = factor(fleet, levels = c("fsh", "bts", "ats"), labels = c("Fishery", "BTS", "ATS"))
  )

# scale bubble sizes within each (fleet x scenario) panel for readability
agecomp_df <- agecomp_df |>
  group_by(fleet, label) |>
  mutate(
    abs_resid = abs(resid),
    # rescale to (0,1] within each panel, but keep a small floor so points always draw
    abs_resid_scaled = pmax(abs_resid / max(abs_resid, na.rm = TRUE), 0.02)
  ) |>
  ungroup()

lim <- max(abs(agecomp_df$resid), na.rm = TRUE)

ggplot(agecomp_df, aes(x = year, y = age)) +
  geom_point(aes(size = abs_resid_scaled, color = resid), alpha = 0.85) +
  facet_grid(fleet ~ label) +
  scale_color_gradient2(low = "#2c7bb6", mid = "white", high = "#d7191c", limits = c(-lim, lim)) +
  scale_size(range = c(0.6, 4.0), guide = "none") +
  labs(x = "Year", y = "Age", color = "Obs-Exp") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: fig-agecomp-fitlines
#| fig-cap: "Observed vs expected age-composition at a few representative ages (proportions), by fleet and scenario."

ages_to_show <- sort(unique(agecomp_df$age))
ages_to_show <- ages_to_show[ages_to_show %in% c(1, 3, 5, 7, 9, 11, 13, 15)]
if (length(ages_to_show) == 0) ages_to_show <- head(sort(unique(agecomp_df$age)), 6)

plot_age_df <- agecomp_df |>
  filter(age %in% ages_to_show)

# expected = black line; observed = points
# facet by fleet x scenario (rows) and age (cols)
ggplot(plot_age_df, aes(x = year)) +
  geom_line(aes(y = exp), linewidth = 0.6, color = "black") +
  geom_point(aes(y = obs), size = 0.5, alpha = 0.7) +
  facet_grid(fleet + label ~ age, scales = "free_y") +
  labs(x = "Year", y = "Proportion", caption = "Black line = expected; points = observed") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The RTMB report object stores a set of likelihood (or penalty) components. The most relevant ones for this comparison are:

- `sel_like`, `sel_like_dev` (selectivity penalties)
- `age_like` (age composition likelihood)
- `cat_like` (catch fit)
- `cpue_like` (CPUE)

```{r}
#| label: tbl-nll-components
#| tbl-cap: "Selected likelihood / penalty components returned in the RTMB report."

extract_components <- function(run) {
  r <- run$report
  tibble(
    fishery_sel_form = run$form,
    label = run$label,
    sel_like = sum(r$sel_like, na.rm = TRUE),
    sel_like_dev = sum(r$sel_like_dev, na.rm = TRUE),
    age_like_total = attr(r$age_like, "total") %||% sum(r$age_like, na.rm = TRUE),
    cat_like = r$cat_like %||% NA_real_,
    cpue_like = r$cpue_like %||% NA_real_,
    avo_like = r$avo_like %||% NA_real_,
    wt_like = r$wt_like %||% NA_real_,
    nll = r$nll %||% NA_real_
  )
}

comp_tbl <- bind_rows(purrr::map(runs, extract_components)) |>
  arrange(fishery_sel_form) |>
  mutate(
    delta_nll = nll - nll[fishery_sel_form == base_form][1],
    delta_sel_like = sel_like - sel_like[fishery_sel_form == base_form][1],
    delta_sel_like_dev = sel_like_dev - sel_like_dev[fishery_sel_form == base_form][1]
  )

comp_tbl |>
  mutate(across(where(is.numeric), ~round(.x, 6))) |>
  gt_report() |>
  tab_header(title = "NLL components")
```

# Fishery Selectivity Surfaces

## Form 5 (2D AR1 time-varying) model statement

For `fishery_sel_form = 5`, the fishery selectivity is constructed from a **year×age latent field** `sel_tv_ar1_fsh[y, a]`.

Let years be indexed by \(y = 1,\dots,n_{yr}\) and ages by \(a = 1,\dots,n_{age}\). Define the (unnormalized) log-selectivity field:

\[
\tilde\ell_{y,a} = u_{y,a}, \quad u_{y,a} \equiv \texttt{sel\_tv\_ar1\_fsh}[y,a].
\]

The model then applies the same centering used elsewhere to enforce an identifiability constraint (roughly, remove an age-average so the scale is anchored):

\[
\ell_{y,\cdot} = \text{center}(\tilde\ell_{y,\cdot})
\]

and the selectivity-at-age is

\[
s_{y,a} = \exp(\ell_{y,a}).
\]

### Which parameters are time-varying?

- **Time-varying:** \(u_{y,a}\) (the full \(n_{yr}\times n_{age}\) matrix `sel_tv_ar1_fsh`). This is what can generate year-to-year changes in selectivity.
- **Static (not time-varying):**
  - `sel_tv_ar1_rho_fsh = (rho_year, rho_age)` (2 parameters controlling AR1 correlation across years and across ages)
  - `log_sel_tv_ar1_sigma_fsh` (1 parameter controlling the marginal scale of the field)
  - `sel_tv_ar1_weight_fsh` (1 scalar multiplier we introduced to relax/strengthen the AR1 penalty; default 1.0)

### Penalty / prior (process model)

The time-varying field is regularized with a **separable AR1×AR1 Gaussian process penalty**:

\[
-\;w\;\log p(u) \;=\; -\;w\;\log\Big( \text{AR1}_y(\rho_y) \otimes \text{AR1}_a(\rho_a) \Big) + \text{const},
\]

implemented as:

- \(\rho_y = \text{logit}^{-1}(\texttt{sel\_tv\_ar1\_rho\_fsh}[1])\)
- \(\rho_a = \text{logit}^{-1}(\texttt{sel\_tv\_ar1\_rho\_fsh}[2])\)
- \(\sigma = \exp(\texttt{log\_sel\_tv\_ar1\_sigma\_fsh})\)
- penalty term: `dseparable(dautoreg, dautoreg)(sel_tv_ar1_fsh, scale = scale)` multiplied by `sel_tv_ar1_weight_fsh = w`.

So if the reported `sel_fsh` still looks nearly constant over time, it means the optimizer is finding that the likelihood gain from moving \(u_{y,a}\) away from 0 is small relative to the AR1 penalty (or the best fit uses only a small amount of variation).

## Selectivity curves (selected years)

```{r}
#| label: fig-sel-curves
#| fig-cap: "Fishery selectivity-at-age curves for selected years, by fishery selectivity form."

years_to_show <- sort(unique(sel_df$year))
years_to_show <- years_to_show[years_to_show %in% c(min(years_to_show), 1980, 1995, 2010, 2024)]
if (length(years_to_show) == 0) years_to_show <- head(sort(unique(sel_df$year)), 5)

plot_df <- sel_df |>
  filter(year %in% years_to_show) |>
  group_by(fishery_sel_form, label, year) |>
  mutate(relative = selectivity / max(selectivity, na.rm = TRUE)) |>
  ungroup()

ggplot(plot_df, aes(x = age, y = relative, color = factor(year))) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~label, ncol = 1, scales = "fixed") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(x = "Age", y = "Relative selectivity", color = "Year")
```

## ggridges (all years)

```{r}
#| label: fig-sel-ridges
#| fig-cap: "Fishery selectivity through time (ggridges), normalized within year."

ridge_df <- sel_df |>
  group_by(fishery_sel_form, label, year) |>
  mutate(relative = selectivity / max(selectivity, na.rm = TRUE)) |>
  ungroup() |>
  mutate(year_factor = factor(year, levels = rev(sort(unique(year)))))

ggplot(ridge_df, aes(x = age, y = year_factor, height = relative, group = year_factor)) +
  geom_density_ridges(stat = "identity", fill = "#1f78b4", alpha = 0.65, rel_min_height = 0.01) +
  facet_wrap(~label, ncol = 3) +
  labs(x = "Age", y = "Year") +
  theme(axis.text.y = element_text(size = 6))
```

# Spawning Biomass Comparison

```{r}
#| label: fig-ssb
#| fig-cap: "Spawning biomass (SSB) trajectories by fishery selectivity form."

ggplot(ssb_df, aes(x = year, y = SSB, color = label)) +
  geom_line(linewidth = 1) +
  labs(x = "Year", y = "SSB", color = NULL)
```

```{r}
#| label: tbl-ssb-differences
#| tbl-cap: "SSB differences relative to base run at key years."

key_years <- sort(unique(ssb_df$year))
key_years <- key_years[key_years %in% c(max(key_years), max(key_years) - 1, 2000, 2010, 2020)]
if (length(key_years) == 0) key_years <- tail(sort(unique(ssb_df$year)), 5)

ssb_wide <- ssb_df |>
  filter(year %in% key_years) |>
  select(fishery_sel_form, label, year, SSB) |>
  pivot_wider(names_from = fishery_sel_form, values_from = SSB, names_prefix = "form_")

base_col <- paste0("form_", base_form)

ssb_delta <- ssb_df |>
  filter(year %in% key_years) |>
  group_by(year) |>
  mutate(
    base_SSB = SSB[fishery_sel_form == base_form][1],
    delta_SSB = SSB - base_SSB,
    pct_delta = 100 * delta_SSB / base_SSB
  ) |>
  ungroup() |>
  select(label, fishery_sel_form, year, SSB, delta_SSB, pct_delta) |>
  arrange(year, fishery_sel_form)

ssb_delta |>
  mutate(across(c(SSB, delta_SSB, pct_delta), ~round(.x, 4))) |>
  gt_report() |>
  tab_header(title = "SSB deltas vs base")
```

# Notes

- If you want, we can add more diagnostics: fishery age-comp fits, residual bubbles, and/or likelihood-by-fleet breakdowns.
