---
title: "Fishery Selectivity Scenario Assessment"
execute:
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    lightbox: true
    code-fold: true
    code-summary: "Show code"
params:
  run_file: "analysis/output/fishery_sel_forms/fishery_sel_form_0.rds"
  scenario_label: "(set via params)"
  run_sparsenuts: false
  force_sparsenuts: false
  sparsenuts_file: "analysis/output/sparsenuts/rtmb_ebswp_sparsenuts_form0.rds"
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggthemes)
  library(ggridges)
  library(gt)
  library(tibble)
  library(readr)
})

param_is_true <- function(x) {
  isTRUE(x) || (
    is.character(x) &&
      length(x) > 0 &&
      tolower(x[1]) %in% c("true", "t", "yes", "y", "1")
  )
}

theme_set(ggthemes::theme_few())

rtmb_root <- normalizePath(if (file.exists("R")) getwd() else file.path(getwd(), ".."), mustWork = TRUE)
rtmb_file <- function(...) normalizePath(file.path(rtmb_root, ...), mustWork = FALSE)

run_path <- rtmb_file(params$run_file)
if (!file.exists(run_path)) stop("Missing run_file: ", run_path)
run <- readRDS(run_path)
rep <- run$report

label_short <- params$scenario_label
if (is.null(label_short) || !nzchar(label_short) || label_short == "(set via params)") {
  label_short <- run$label %||% paste0("form ", run$form)
}

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0) y else x

as_year_age <- function(mat, years, ages = seq_len(ncol(mat)), value_name = "value") {
  out <- as_tibble(mat, .name_repair = "minimal")
  names(out) <- ages
  out |>
    mutate(year = years, .before = 1) |>
    pivot_longer(-year, names_to = "age", values_to = value_name) |>
    mutate(age = as.integer(age))
}
```

# Scenario

- File: `{params$run_file}`
- Label: `{label_short}`

# Fit summary

```{r}
#| label: tbl-fit-summary
#| tbl-cap: "Optimization summary (from saved run file)."

tibble(
  fishery_sel_form = run$form,
  label = run$label,
  objective = run$objective,
  convergence = run$convergence,
  max_gradient = run$max_gradient,
  seconds = run$seconds
) |>
  gt()
```

# Catch biomass fit

```{r}
#| label: fig-catch-fit
#| fig-cap: "Observed vs predicted catch biomass."
obs <- rep$obs_catch
pred <- rep$pred_catch
stopifnot(!is.null(obs), !is.null(pred))
yrs <- as.integer(names(obs) %||% names(pred) %||% seq_along(obs))
df <- tibble(year = yrs, obs = as.numeric(obs), pred = as.numeric(pred))

ggplot(df, aes(x = year)) +
  geom_line(aes(y = obs), linewidth = 0.6) +
  geom_line(aes(y = pred), linewidth = 0.6, linetype = 2) +
  labs(x = "Year", y = "Catch biomass", caption = "Solid=Observed; Dashed=Predicted")
```

# SSB

```{r}
#| label: fig-ssb
#| fig-cap: "SSB time series."
ssb <- rep$SSB
ssb_df <- tibble(year = as.integer(names(ssb)), SSB = as.numeric(ssb))

ggplot(ssb_df, aes(year, SSB)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Year", y = "SSB")
```

# Fishery selectivity

```{r}
#| label: fig-sel-surface
#| fig-cap: "Fishery selectivity surface (relative within year)."
sel <- rep$sel_fsh
stopifnot(is.matrix(sel))
years <- as.integer(rownames(sel) %||% seq_len(nrow(sel)))
sel_df <- as_year_age(sel, years, value_name = "sel") |>
  group_by(year) |>
  mutate(sel_rel = sel / max(sel, na.rm = TRUE)) |>
  ungroup()

ggplot(sel_df, aes(x = year, y = age, fill = sel_rel)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "Year", y = "Age", fill = "Rel sel")
```

# Age compositions (Obs vs Exp, within-year sum-to-1)

```{r}
#| label: fig-agecomp-resid
#| fig-cap: "Age-comp residuals (Obs - Exp) bubble plot, by fleet."

extract_agecomp <- function(fleet) {
  oac <- rep[[paste0("oac_", fleet)]]
  eac <- rep[[paste0("phat_", fleet)]]
  yrs <- rep[[paste0("yrs_", fleet, "_data")]]
  stopifnot(!is.null(oac), !is.null(eac), !is.null(yrs))

  ages <- seq_len(ncol(oac))
  # normalize within-year over ages
  oac <- sweep(oac, 1, pmax(rowSums(oac), 1e-12), "/")
  eac <- sweep(eac, 1, pmax(rowSums(eac), 1e-12), "/")

  df <- tidyr::expand_grid(year = yrs, age = ages) |>
    mutate(
      obs = as.numeric(oac),
      exp = as.numeric(eac),
      resid = obs - exp,
      fleet = fleet
    )

  # drop age1 for BTS/ATS for display consistency with prior diagnostics
  if (fleet %in% c("bts", "ats")) df <- df |> filter(age >= 2)
  df
}

df <- bind_rows(extract_agecomp("fsh"), extract_agecomp("bts"), extract_agecomp("ats")) |>
  mutate(fleet = factor(fleet, levels = c("fsh", "bts", "ats"), labels = c("Fishery", "BTS", "ATS")))

lim <- max(abs(df$resid), na.rm = TRUE)

ggplot(df, aes(x = year, y = age)) +
  geom_point(aes(size = abs(resid), color = resid), alpha = 0.8) +
  facet_wrap(~fleet, ncol = 1) +
  scale_color_gradient2(low = "#2c7bb6", mid = "white", high = "#d7191c", limits = c(-lim, lim)) +
  scale_size(range = c(0.2, 3.5), guide = "none") +
  labs(x = "Year", y = "Age", color = "Obs-Exp")
```

# SparseNUTS MCMC (slow-mixing marginals + pairs)

This section uses `SparseNUTS` to sample the RTMB posterior and then produces:

- **Marginal plots** for the slowest-mixing parameters (by ESS)
- **Pairs plot** for the same slow parameters

It will **reuse** a saved run if `params$sparsenuts_file` exists, unless you render with `force_sparsenuts: true`.

```{r}
#| label: sparsenuts-run
#| echo: true

sparsenuts_available <- requireNamespace("SparseNUTS", quietly = TRUE)

sparsenuts_file <- rtmb_file(params$sparsenuts_file)

sparsenuts_fit <- NULL
sparsenuts_error <- NULL

if (file.exists(sparsenuts_file) && !param_is_true(params$force_sparsenuts)) {
  sparsenuts_fit <- readRDS(sparsenuts_file)
} else if (param_is_true(params$run_sparsenuts)) {
  if (!sparsenuts_available) {
    sparsenuts_error <- "SparseNUTS is not installed."
  } else {
    dir.create(dirname(sparsenuts_file), showWarnings = FALSE, recursive = TRUE)

    # NOTE: The saved scenario run files do not include the RTMB obj.
    # We rebuild the RTMB object for this scenario by sourcing config.R,
    # setting fishery_sel_form, and refitting at the saved MLE.

    pollock_root <- Sys.getenv("POLLOCK_ROOT", unset = NA_character_)
    if (is.na(pollock_root) || !nzchar(pollock_root)) pollock_root <- Sys.getenv("POLLOCK_BASE", unset = NA_character_)
    if (is.na(pollock_root) || !nzchar(pollock_root)) stop("Set POLLOCK_ROOT before running SparseNUTS.")

    rtmb_env <- new.env(parent = globalenv())
    rtmb_env$rm <- function(...) invisible(NULL)
    rtmb_env$source <- function(file, ...) base::source(file, local = parent.frame(), ...)
    base::source(rtmb_file("R", "config.R"), local = rtmb_env)

    data <- rtmb_env$data
    parms <- rtmb_env$parms
    rpm <- rtmb_env$rpm

    # Override scenario selectivity form and rebuild mapping so the right parameter blocks are active
    data$fishery_sel_form <- as.integer(run$form)
    parms$fishery_sel_form <- as.integer(run$form)

    # Always ensure optional blocks exist with correct dimensions
    parms <- add_fishery_selectivity_parameters(parms, data)

    fixed_params <- c(
      "log_K", "d_scale", "L1", "L2",
      "log_avginit", "log_avgrec", "log_avg_F",
      "log_q_bts",
      "rec_dev_future",
      "natmort_phi",
      "larv_rec_devs",
      "sel_devs_bts", "sel_slp_bts", "sel_a50_bts", "sel_age_one_bts",
      "sel_trm2_fsh", "sel_dif1_fsh", "sel_a501_fsh", "sel_trm1_fsh",
      "sel_dif2_fsh", "sel_dif1_fsh_dev", "sel_a501_fsh_dev", "sel_trm2_fsh_dev",
      inactive_fishery_selectivity_parameters(data$fishery_sel_form),
      "resid_temp_x1", "resid_temp_x2",
      "log_q_std_area", "bt_slope", "sigr", "steepness",
      "sel_coffs_bts",
      "M_dev",
      "log_a_II", "log_b_II", "log_a_II_vec", "log_b_II_vec",
      "log_rho", "log_resid_M",
      "log_alpha"
    )

    map_obj <- create_map_from_par(
      parms, parms,
      exact_names = fixed_params,
      exclude_patterns = "xxx"
    )

    obj <- RTMB::MakeADFun(rpm, parms, map = map_obj)

    # initialize at saved MLE if available
    mle_par <- run$par %||% NULL
    if (!is.null(mle_par)) {
      obj$par[names(mle_par)] <- unlist(mle_par)
    }

    sparsenuts_fit <- tryCatch(
      SparseNUTS::sample_snuts(
        obj,
        cores = 1,
        globals = list(data = data)
      ),
      error = function(e) {
        sparsenuts_error <<- conditionMessage(e)
        NULL
      }
    )

    if (!is.null(sparsenuts_fit)) saveRDS(sparsenuts_fit, sparsenuts_file)
  }
}

list(
  sparsenuts_available = sparsenuts_available,
  run_sparsenuts = param_is_true(params$run_sparsenuts),
  force_sparsenuts = param_is_true(params$force_sparsenuts),
  sparsenuts_file = sparsenuts_file,
  has_fit = !is.null(sparsenuts_fit),
  error = sparsenuts_error
)
```

```{r}
#| label: sparsenuts-plots
#| echo: false
#| fig-cap: "SparseNUTS diagnostics for slow-mixing parameters (marginals + pairs)."

if (!is.null(sparsenuts_fit) && sparsenuts_available && !is.null(sparsenuts_fit$samples) && !is.null(sparsenuts_fit$monitor)) {

  fig_dir <- rtmb_file("analysis", "output", "figures")
  dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

  mon <- sparsenuts_fit$monitor
  mon <- mon |> arrange(ess_bulk)
  slow <- head(mon$variable, 12)
  slow <- slow[slow %in% dimnames(sparsenuts_fit$samples)[[3]]]

  pairs_file <- file.path(fig_dir, sprintf("sparsenuts_pairs_slow_form%d.png", as.integer(run$form)))
  marg_file <- file.path(fig_dir, sprintf("sparsenuts_marginals_slow_form%d.png", as.integer(run$form)))

  # Pairs plot
  png(pairs_file, width = 1800, height = 1400, res = 180)
  try(SparseNUTS::plot_pairs(sparsenuts_fit, pars = slow, max_points = 2000), silent = TRUE)
  dev.off()

  # Marginals plot
  png(marg_file, width = 1800, height = 1400, res = 180)
  try(SparseNUTS::plot_marginals(sparsenuts_fit, pars = slow, ncol = 3), silent = TRUE)
  dev.off()

  knitr::include_graphics(c(pairs_file, marg_file))

} else {
  cat("No SparseNUTS fit available. Render with run_sparsenuts: true to run / load MCMC and produce plots.\n")
}
```
