---
title: "RTMBprof Case Study: Acoustic-survey Catchability in the EBS Pollock Model"
subtitle: "Constrained component profiles and posterior-objective evaluations for log q~ATS~"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    lightbox: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(RTMBprof)
  library(ggplot2)
  library(knitr)
  library(plotly)
})

theme_set(theme_minimal(base_size = 12))

repo_root <- normalizePath(
  if (file.exists("R")) "." else "..",
  mustWork = TRUE
)
stage1_path <- file.path(
  repo_root, "analysis", "output", "rtmbprof_stage1",
  "candidate_scale_profiles.rds"
)
stage2_path <- file.path(
  repo_root, "analysis", "output", "rtmbprof_stage2",
  "rtmbprof_stage2.rds"
)
if (!file.exists(stage1_path)) {
  stop("Missing Stage 1 output. Run R/run_rtmbprof_stage1.R.")
}
if (!file.exists(stage2_path)) {
  stop("Missing Stage 2 output. Run R/run_rtmbprof_stage2.R.")
}

stage1 <- readRDS(stage1_path)
stage2 <- readRDS(stage2_path)
profile <- stage1$profiles$log_q_ats
mcmc_components <- stage2$mcmc_components
mcmc_summary <- stage2$mcmc_summary
profile_components <- attr(profile, "components")

fitted_index <- which.min(profile$value)
fitted_q <- profile$log_q_ats[fitted_index]
delta_nll <- profile$value - min(profile$value)
left_crossed <- any(delta_nll[profile$log_q_ats < fitted_q] >= 1.92)
right_crossed <- any(delta_nll[profile$log_q_ats > fitted_q] >= 1.92)

fmt <- function(x, digits = 3) {
  formatC(x, digits = digits, format = "fg")
}
```

## Overview {#sec-overview}

This case study applies `RTMBprof` to the RTMB implementation of the eastern
Bering Sea pollock assessment. The focal parameter is `log_q_ats`, the log
catchability for the acoustic-trawl survey (ATS). It was selected because
candidate profiles showed broader leverage on population scale than
`log_q_cpue`. The effective BTS catchability cannot be profiled in the current
model because `log_q_bts` is fixed and `q_bts` is calculated analytically.

The analysis has two deliberately distinct parts:

1. a **constrained joint-objective profile**, which fixes `log_q_ats` and
   reoptimizes all nuisance parameters; and
2. **posterior-objective evaluations**, which retain the nuisance-parameter
   values at each saved SparseNUTS draw.

The second calculation is not another likelihood profile. Section
@sec-profile describes the constrained profile, @sec-components identifies
which objective terms create its shape, and @sec-posterior compares it with
the posterior evaluations.

## Model and analysis configuration {#sec-methods}

The fitted RTMB object has no parameters designated as Laplace random effects,
so both stages evaluate the complete **joint** negative log posterior. At every
profile point and posterior draw, the reported components must sum to the
objective within numerical tolerance.

```{r}
#| label: tbl-analysis
#| tbl-cap: "Configuration and numerical checks for the qATS case study."

analysis_table <- data.frame(
  Quantity = c(
    "Profile target",
    "Constrained profile points",
    "Profile objective",
    "Profile threshold crossed",
    "Maximum profile closure error",
    "Posterior chains",
    "Warmup removed per chain",
    "Retained posterior draws",
    "Maximum posterior closure error"
  ),
  Value = c(
    "log_q_ats",
    nrow(profile),
    attr(profile, "objective"),
    if (left_crossed && right_crossed) "Both sides of ΔNLL = 1.92" else "No",
    format(max(abs(profile$closure_error)), scientific = TRUE, digits = 3),
    stage2$diagnostics$chains,
    stage2$diagnostics$warmup_removed_per_chain,
    stage2$diagnostics$retained_draws,
    format(
      stage2$diagnostics$max_abs_closure_error,
      scientific = TRUE,
      digits = 3
    )
  ),
  check.names = FALSE
)
kable(analysis_table, align = c("l", "l"), row.names = FALSE)
```

As shown in @tbl-analysis, the component closure checks pass at all 27
constrained points and all 4,000 retained posterior draws.

## Constrained qATS profile {#sec-profile}

The profile spans `r fmt(min(profile$log_q_ats))` to
`r fmt(max(profile$log_q_ats))` on the log scale and crosses the conventional
one-parameter 95% objective cutoff, $\Delta\mathrm{NLL}=1.92$, on both sides
of its minimum.

```{r}
#| label: fig-total-profile
#| fig-cap: "Constrained joint-objective profile for log acoustic-trawl survey catchability. The horizontal line marks ΔNLL = 1.92 and the vertical dotted line marks the profile minimum."
#| fig-width: 8
#| fig-height: 5

profile_plot_data <- data.frame(
  log_q_ats = profile$log_q_ats,
  delta_nll = delta_nll
)
ggplot(profile_plot_data, aes(log_q_ats, delta_nll)) +
  geom_hline(yintercept = 1.92, linetype = "dashed", colour = "#D55E00") +
  geom_vline(xintercept = fitted_q, linetype = "dotted", colour = "grey35") +
  geom_line(linewidth = 1, colour = "#0072B2") +
  geom_point(size = 1.7, colour = "#0072B2") +
  labs(
    x = "log acoustic-trawl survey catchability",
    y = expression(Delta * " joint negative log posterior")
  )
```

@Fig-total-profile establishes that `log_q_ats` is estimable over the explored
range. The response of biomass-related quantities along that profile is
summarized in @tbl-scale-response.

```{r}
#| label: tbl-scale-response
#| tbl-cap: "Changes in population-scale quantities across the constrained qATS profile."

scale_data <- stage1$scale_diagnostics$log_q_ats
scale_table <- data.frame(
  Quantity = c(
    "Terminal spawning biomass",
    "Mean spawning biomass",
    "Unfished spawning biomass (Bzero)",
    "Analytically derived BTS catchability"
  ),
  Minimum = c(
    min(scale_data$terminal_SSB),
    min(scale_data$mean_SSB),
    min(scale_data$Bzero),
    min(scale_data$q_bts)
  ),
  Maximum = c(
    max(scale_data$terminal_SSB),
    max(scale_data$mean_SSB),
    max(scale_data$Bzero),
    max(scale_data$q_bts)
  )
)
scale_table$`Range change (%)` <-
  100 * (scale_table$Maximum / scale_table$Minimum - 1)
kable(
  scale_table,
  digits = c(NA, 3, 3, 2),
  align = c("l", "r", "r", "r"),
  row.names = FALSE
)
```

## Objective-component profiles {#sec-components}

The complete component decomposition in @fig-component-profile shows how
individual data sources and penalties respond while nuisance parameters are
reoptimized. Curves are centered separately at the profile minimum so their
changes, rather than their absolute offsets, are comparable.

```{r}
#| label: fig-component-profile
#| fig-cap: "Constrained qATS profile decomposed into all reported objective components. Every curve is expressed relative to its value at the profile minimum; the heavier blue curve is the total joint objective."
#| fig-width: 10
#| fig-height: 7

component_columns <- c("value", profile_components)
component_labels <- c("Total", profile_components)
component_plot_data <- do.call(
  rbind,
  lapply(seq_along(component_columns), function(i) {
    values <- profile[[component_columns[i]]]
    data.frame(
      log_q_ats = profile$log_q_ats,
      component = component_labels[i],
      change = values - values[fitted_index]
    )
  })
)
component_plot_data$component <- factor(
  component_plot_data$component,
  levels = component_labels
)
component_plot_data$hover <- sprintf(
  paste(
    "<b>%s</b>",
    "log q<sub>ATS</sub>: %.4f",
    "Change: %.4f",
    sep = "<br>"
  ),
  component_plot_data$component,
  component_plot_data$log_q_ats,
  component_plot_data$change
)

component_ggplot <- ggplot(
  component_plot_data,
  aes(
    log_q_ats,
    change,
    colour = component,
    group = component,
    linewidth = component,
    text = hover
  )
) +
  geom_line() +
  scale_linewidth_manual(
    values = c(Total = 1.35, setNames(rep(0.65, length(profile_components)),
                                     profile_components)),
    guide = "none"
  ) +
  labs(
    x = "log acoustic-trawl survey catchability",
    y = "Change from value at profile minimum",
    colour = "Component"
  ) +
  theme(legend.position = "right")

ggplotly(component_ggplot, tooltip = "text") |>
  layout(
    legend = list(
      orientation = "v",
      x = 1.02,
      y = 1
    ),
    margin = list(r = 180)
  )
```

To rank component movement without conflating direction, @tbl-component-span
reports each component's range over the profile.

```{r}
#| label: tbl-component-span
#| tbl-cap: "Objective-component ranges across the constrained qATS profile, ordered from largest to smallest."

component_span <- data.frame(
  Component = profile_components,
  `Objective range` = vapply(
    profile[profile_components],
    function(x) diff(range(x)),
    numeric(1)
  ),
  check.names = FALSE
)
component_span <- component_span[
  order(component_span$`Objective range`, decreasing = TRUE),
  ,
  drop = FALSE
]
kable(component_span, digits = 4, align = c("l", "r"), row.names = FALSE)
```

The age-composition and BTS terms have the largest ranges, followed by ATS
itself. This is a model-wide scale adjustment rather than an isolated response
of the focal survey likelihood.

## Posterior-objective comparison {#sec-posterior}

The saved SparseNUTS sample has strong diagnostics for the focal parameter
(@tbl-mcmc-diagnostics). Warmup was removed explicitly before passing the
iterations-by-chains-by-variables array to `RTMBprof`.

```{r}
#| label: tbl-mcmc-diagnostics
#| tbl-cap: "SparseNUTS and RTMBprof posterior-evaluation diagnostics."

d <- stage2$diagnostics
mcmc_diagnostic_table <- data.frame(
  Diagnostic = c(
    "log_q_ats R-hat",
    "log_q_ats bulk ESS",
    "log_q_ats tail ESS",
    "Fitted dimension",
    "Gaussian expected objective gap (d/2)",
    "Observed median objective gap",
    "Maximum component closure error"
  ),
  Value = c(
    fmt(d$log_q_ats_rhat, 6),
    fmt(d$log_q_ats_ess_bulk, 6),
    fmt(d$log_q_ats_ess_tail, 6),
    d$fitted_dimension,
    fmt(d$gaussian_expected_objective_gap),
    fmt(d$median_objective_gap_from_profile),
    format(d$max_abs_closure_error, scientific = TRUE, digits = 3)
  )
)
kable(mcmc_diagnostic_table, align = c("l", "r"), row.names = FALSE)
```

In a locally Gaussian posterior of dimension $d$, a typical draw lies about
$d/2$ objective units above the mode. The observed median gap of
`r fmt(d$median_objective_gap_from_profile)` closely matches the
`r fmt(d$gaussian_expected_objective_gap)`-unit expectation. Consequently, the
raw posterior cloud in @fig-profile-posterior is vertically separated from the
constrained optimum; this separation is expected geometry, not failed
profiling.

```{r}
#| label: fig-profile-posterior
#| fig-cap: "Objective values at individual posterior draws (grey), their binned conditional median (orange), and the constrained joint-objective profile (blue). Posterior draws retain rather than reoptimize nuisance parameters."
#| fig-width: 9
#| fig-height: 5.5

plot(
  mcmc_components,
  profile = profile,
  summary = mcmc_summary
) +
  labs(
    x = "log acoustic-trawl survey catchability",
    y = "Change in joint negative log posterior"
  )
```

Separately centering the constrained profile and posterior evaluations exposes
their one-dimensional shapes without presenting the orange curve as a
likelihood profile (@fig-centred-shape).

```{r}
#| label: fig-centred-shape
#| fig-cap: "Separately centered qATS shape diagnostic. The orange curve and grey interval are binned posterior-objective summaries, whereas the blue curve is the constrained joint-objective profile."
#| fig-width: 9
#| fig-height: 5.5

plot_mcmc_shape(
  mcmc_components,
  profile = profile,
  summary = mcmc_summary
) +
  labs(x = "log acoustic-trawl survey catchability")
```

Selected component-level comparisons are shown in
@fig-component-posterior. The profile curves optimize nuisance parameters at
each focal value, whereas each grey point preserves a complete posterior
parameter vector.

```{r}
#| label: fig-component-posterior
#| fig-cap: "Posterior evaluations, binned 95% intervals and medians, and constrained profiles for the total objective and the six components with the largest constrained-profile ranges."
#| fig-width: 10
#| fig-height: 11

display_components <- c(
  "value",
  head(component_span$Component, 6)
)
plot(
  mcmc_summary,
  component = display_components,
  profile = profile,
  evaluations = mcmc_components,
  ncol = 2
) +
  labs(x = "log acoustic-trawl survey catchability")
```

## Parameter associations {#sec-associations}

The posterior associations in @tbl-correlations support the original choice
of scale parameter. `log_q_ats` is substantially associated with the AVO
catchability, but nearly uncorrelated with `log_Rzero`. Thus, in this fitted
model, `log_Rzero` primarily describes stock--recruitment behavior rather than
the principal survey-scale trade-off.

```{r}
#| label: tbl-correlations
#| tbl-cap: "Posterior correlations of selected parameters with log qATS."

correlation_table <- stage2$posterior_correlations
names(correlation_table) <- c("Parameter", "Correlation with log_q_ats")
kable(
  correlation_table,
  digits = 3,
  align = c("l", "r"),
  row.names = FALSE
)
```

## Conclusions {#sec-conclusions}

1. `log_q_ats` produces a well-resolved constrained profile that crosses the
   one-parameter 95% objective cutoff on both sides.
2. Its profile changes terminal spawning biomass by approximately
   `r fmt(100 * (max(scale_data$terminal_SSB) /
     min(scale_data$terminal_SSB) - 1), 3)`%, demonstrating meaningful
   population-scale leverage.
3. Age-composition, BTS, and ATS components contribute most strongly to the
   profile shape.
4. All component sums close against the joint objective at numerical
   precision for both constrained points and posterior draws.
5. The constrained profile and binned posterior-objective summaries answer
   different questions. Their comparison is informative only when that
   distinction—and the expected high-dimensional objective offset—is retained.

## Reproduction {#sec-reproduction}

From the repository root:

```bash
Rscript R/run_rtmbprof_stage1.R
Rscript R/run_rtmbprof_stage2.R
quarto render reporting/rtmbprof_q_ats_case_study.qmd
```

The Quarto document reads the resulting RDS files and does not refit the model
or rerun SparseNUTS during rendering.
