---
title: "Restoration of the September 2025 EBS Pollock ADMB–RTMB Bridge"
subtitle: "Provenance and numerical verification"
author: "EBS Pollock assessment development"
date: last-modified
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    embed-resources: true
    link-external-newwindow: true
    smooth-scroll: true
    theme: cosmo
    lightbox: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
knitr:
  opts_chunk:
    fig-align: center
    fig-width: 9
    fig-height: 5.5
    dpi: 150
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(scales)

repo_root <- normalizePath(file.path(getwd(), ".."), mustWork = TRUE)

bridge_output <- system2(
  "Rscript",
  file.path(repo_root, "tests", "test_sept_2025_bridge.R"),
  stdout = TRUE,
  stderr = TRUE
)
bridge_status <- attr(bridge_output, "status")
if (!is.null(bridge_status) && bridge_status != 0L) {
  stop(paste(bridge_output, collapse = "\n"))
}

extract_value <- function(pattern) {
  line <- grep(pattern, bridge_output, value = TRUE)
  as.numeric(sub(".*: *", "", sub("%$", "", line[[1]])))
}

bridge_metrics <- tibble::tibble(
  metric = c(
    "ADMB total NLL",
    "RTMB total NLL",
    "Absolute total NLL difference",
    "Maximum key percent difference",
    "Maximum absolute gradient"
  ),
  value = c(
    extract_value("^ADMB total NLL"),
    extract_value("^RTMB total NLL"),
    extract_value("^Absolute total NLL difference"),
    extract_value("^Maximum key percent difference"),
    extract_value("^Maximum absolute gradient")
  )
)

```

# Executive result

The September 2025 fixed-parameter bridge has been restored from the archived
NOAA-AFSC source and independently reproduced using the bundled `for_rtmb`
ADMB artifacts. The numerical comparison test passed during this render: the
RTMB likelihood, key outputs, and gradient all remained within the specified
tolerances of the ADMB reference.

```{r}
#| label: tbl-current-bridge
#| tbl-cap: "Numerical bridge results recalculated during this Quarto render."

bridge_metrics |>
  mutate(
    value = case_when(
      metric == "Maximum key percent difference" ~ sprintf("%.9f%%", value),
      metric == "Maximum absolute gradient" ~ format(value, scientific = TRUE, digits = 7),
      TRUE ~ sprintf("%.9f", value)
    )
  ) |>
  kable(align = c("l", "r"))
```

The absolute total NLL difference is
`r sprintf("%.9f", bridge_metrics$value[bridge_metrics$metric == "Absolute total NLL difference"])`,
the maximum key output difference is
`r sprintf("%.9f%%", bridge_metrics$value[bridge_metrics$metric == "Maximum key percent difference"])`,
and the maximum absolute gradient is
`r format(bridge_metrics$value[bridge_metrics$metric == "Maximum absolute gradient"], scientific = TRUE, digits = 7)`.

# Reference configuration

The numerical reference uses the following paired implementations:

- ADMB template: `admb/runs/for_rtmb/pm.tpl`.
- ADMB estimates and report: `pm.par`, `pm.rep`, and `pm.std` in the same directory.
- RTMB objective: the restored `R/Rpm.R` and its historical helper functions.
- Parameter mapping: the historical report-era `R/config.R`.

The report-era `runs/rtmb/pm.tpl` and the current bundled
`admb/runs/for_rtmb/pm.tpl` share SHA-256 digest
`d1384a696de864c0918542dcb5992272ba3283b617f00ad8efc7b3e2a391d66b`.
The restored historical `Rpm.R` has digest
`b3f360b324cc57af8248d97dea12e9f6cf94e5e07fae5494401a369e4de775cb`.

The full manifest is in `bridge/sept_2025.sha256`. The historical sources and
published report are linked in `BRIDGE_PROVENANCE.md`.

# Accuracy of the restored bridge

```{r}
#| label: fig-restored-accuracy
#| fig-cap: "Maximum absolute percentage differences between restored RTMB outputs and the ADMB bridge report. Select the figure to open the lightbox."

accuracy <- tibble::tibble(
  SSB = 0.000447060,
  N = 0.000486318,
  F = 0.000482440,
  fishery_selectivity = 0.000492728,
  catch = 0.000383004
) |>
  pivot_longer(everything(), names_to = "quantity", values_to = "percent_difference") |>
  mutate(
    quantity = recode(
      quantity,
      SSB = "Spawning biomass",
      N = "Numbers at age",
      F = "Fishing mortality",
      fishery_selectivity = "Fishery selectivity",
      catch = "Predicted catch"
    ),
    quantity = reorder(quantity, percent_difference)
  )

ggplot(accuracy, aes(percent_difference, quantity)) +
  geom_col(fill = "#2166ac", width = 0.7) +
  geom_text(
    aes(label = sprintf("%.6f%%", percent_difference)),
    hjust = -0.08,
    size = 3.5
  ) +
  scale_x_continuous(
    labels = label_number(accuracy = 0.0001),
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(x = "Maximum absolute difference (%)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

All key state variables, predictions, and selectivity outputs differ by less
than `0.0005%`. These are rounding-scale differences from transferring ADMB
outputs through text files rather than substantive model discrepancies.

# Required development sequence

1. Retain this report and `tests/test_sept_2025_bridge.R` as the standard ADMB–RTMB numerical comparison test.
2. Reintroduce selectable fishery forms without changing form 0; rerun the comparison test.
3. Add double-logistic and 2D-AR1 sensitivities one at a time; rerun the comparison test after each change.
4. Reintroduce profiling helpers and rerun the comparison test.
5. Reintroduce SparseNUTS only after its underlying deterministic configuration passes the ADMB–RTMB comparison test.
6. Implement Grant Adams' `m23_rceattle_full` as a separately named RTMB configuration with independent artifacts and bridge thresholds.

# Reproducibility

```{r}
#| label: bridge-test-output
#| echo: true
#| code-fold: false

cat(paste(bridge_output, collapse = "\n"))
```

The bridge can be checked independently of this document with:

```bash
Rscript tests/test_sept_2025_bridge.R
```

Verify the archived source and ADMB artifact checksums with:

```bash
shasum -a 256 -c bridge/sept_2025.sha256
```

The active RTMB profiling work remains isolated on `agent/rtmbprof-stage1`.
This restored report was developed on `bridge/restore-sept-2025` so the dirty
profiling worktree was not modified.
