---
title: "Restoration of the September 2025 EBS Pollock ADMB–RTMB Bridge"
subtitle: "Provenance, numerical verification, and post-bridge regression audit"
author: "EBS Pollock assessment development"
date: last-modified
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    embed-resources: true
    link-external-newwindow: true
    smooth-scroll: true
    theme: cosmo
    lightbox: true
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
knitr:
  opts_chunk:
    fig-align: center
    fig-width: 9
    fig-height: 5.5
    dpi: 150
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(scales)

repo_root <- normalizePath(file.path(getwd(), ".."), mustWork = TRUE)

bridge_output <- system2(
  "Rscript",
  file.path(repo_root, "tests", "test_sept_2025_bridge.R"),
  stdout = TRUE,
  stderr = TRUE
)
bridge_status <- attr(bridge_output, "status")
if (!is.null(bridge_status) && bridge_status != 0L) {
  stop(paste(bridge_output, collapse = "\n"))
}

extract_value <- function(pattern) {
  line <- grep(pattern, bridge_output, value = TRUE)
  as.numeric(sub(".*: *", "", sub("%$", "", line[[1]])))
}

bridge_metrics <- tibble::tibble(
  metric = c(
    "ADMB total NLL",
    "RTMB total NLL",
    "Absolute total NLL difference",
    "Maximum key percent difference",
    "Maximum absolute gradient"
  ),
  value = c(
    extract_value("^ADMB total NLL"),
    extract_value("^RTMB total NLL"),
    extract_value("^Absolute total NLL difference"),
    extract_value("^Maximum key percent difference"),
    extract_value("^Maximum absolute gradient")
  )
)

audit <- tibble::tribble(
  ~source, ~nll_difference, ~gradient, ~SSB, ~N, ~F, ~fishery_selectivity, ~catch,
  "September 2025 archived bridge", -0.004097748, 0.00000636535, 0.000447060, 0.000486318, 0.000482440, 0.000492728, 0.000383004,
  "NOAA-AFSC 215efa4",             -0.004097748, 0.00001490000, 0.000447060, 0.000486318, 0.000482440, 0.000492728, 0.000383004,
  "Standalone 567e2ee",              0.506449000, 4.68066000000, 0.000447060, 0.000486318, 0.000482440, 0.000492728, 0.000383004,
  "Selectable forms 7301aa0",        0.506449000, 4.68066000000, 0.000447060, 0.000486318, 0.000482440, 0.000492728, 0.000383004,
  "Pre-cap af5c490",                 0.506449000, 4.68066000000, 0.000447060, 0.000486318, 0.000482440, 0.000492728, 0.000383004,
  "Old-age cap 7db7217",             10.73690000, 113.409000000, 3.479380000, 12.05650000, 26.53340000, 26.53360000, 6.519860000,
  "Later comparison cd46353",        10.73690000, 113.409000000, 3.479380000, 12.05650000, 26.53340000, 26.53360000, 6.519860000
)
```

# Executive result

The September 2025 fixed-parameter bridge has been restored from the archived
NOAA-AFSC source and independently reproduced using the bundled `for_rtmb`
ADMB artifacts. The automated regression gate passed during this render.

```{r}
#| label: tbl-current-bridge
#| tbl-cap: "Numerical bridge results recalculated during this Quarto render."

bridge_metrics |>
  mutate(
    value = case_when(
      metric == "Maximum key percent difference" ~ sprintf("%.9f%%", value),
      metric == "Maximum absolute gradient" ~ format(value, scientific = TRUE, digits = 7),
      TRUE ~ sprintf("%.9f", value)
    )
  ) |>
  kable(align = c("l", "r"))
```

The absolute total NLL difference is
`r sprintf("%.9f", bridge_metrics$value[bridge_metrics$metric == "Absolute total NLL difference"])`,
the maximum key output difference is
`r sprintf("%.9f%%", bridge_metrics$value[bridge_metrics$metric == "Maximum key percent difference"])`,
and the maximum absolute gradient is
`r format(bridge_metrics$value[bridge_metrics$metric == "Maximum absolute gradient"], scientific = TRUE, digits = 7)`.

# Reference configuration

The numerical reference uses the following paired implementations:

- ADMB template: `admb/runs/for_rtmb/pm.tpl`.
- ADMB estimates and report: `pm.par`, `pm.rep`, and `pm.std` in the same directory.
- RTMB objective: the restored `R/Rpm.R` and its historical helper functions.
- Parameter mapping: the historical report-era `R/config.R`.

The report-era `runs/rtmb/pm.tpl` and the current bundled
`admb/runs/for_rtmb/pm.tpl` share SHA-256 digest
`d1384a696de864c0918542dcb5992272ba3283b617f00ad8efc7b3e2a391d66b`.
The restored historical `Rpm.R` has digest
`b3f360b324cc57af8248d97dea12e9f6cf94e5e07fae5494401a369e4de775cb`.

The full manifest is in `bridge/sept_2025.sha256`. The historical sources and
published report are linked in `BRIDGE_PROVENANCE.md`.

# Accuracy of the restored bridge

```{r}
#| label: fig-restored-accuracy
#| fig-cap: "Maximum absolute percentage differences between restored RTMB outputs and the ADMB bridge report. Select the figure to open the lightbox."

accuracy <- audit |>
  slice(1) |>
  select(SSB, N, F, fishery_selectivity, catch) |>
  pivot_longer(everything(), names_to = "quantity", values_to = "percent_difference") |>
  mutate(
    quantity = recode(
      quantity,
      SSB = "Spawning biomass",
      N = "Numbers at age",
      F = "Fishing mortality",
      fishery_selectivity = "Fishery selectivity",
      catch = "Predicted catch"
    ),
    quantity = reorder(quantity, percent_difference)
  )

ggplot(accuracy, aes(percent_difference, quantity)) +
  geom_col(fill = "#2166ac", width = 0.7) +
  geom_text(
    aes(label = sprintf("%.6f%%", percent_difference)),
    hjust = -0.08,
    size = 3.5
  ) +
  scale_x_continuous(
    labels = label_number(accuracy = 0.0001),
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(x = "Maximum absolute difference (%)", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

All key state variables, predictions, and selectivity outputs differ by less
than `0.0005%`. These are rounding-scale differences from transferring ADMB
outputs through text files rather than substantive model discrepancies.

# Post-bridge regression audit

```{r}
#| label: tbl-regression-audit
#| tbl-cap: "Numerical audit of selected source revisions. State and prediction columns are maximum absolute percentage differences from ADMB."

audit |>
  mutate(across(where(is.numeric), ~ format(.x, scientific = FALSE, digits = 7))) |>
  kable(
    col.names = c(
      "Source state", "NLL difference", "Maximum gradient", "SSB %", "N %",
      "F %", "Fishery selectivity %", "Catch %"
    ),
    align = c("l", rep("r", 7))
  )
```

```{r}
#| label: fig-regression-timeline
#| fig-cap: "Growth of state and prediction discrepancies across selected post-bridge source states. The logarithmic scale separates rounding-level bridge differences from the old-age selectivity regression. Select the figure to open the lightbox."

audit |>
  select(source, SSB, N, F, fishery_selectivity, catch) |>
  pivot_longer(-source, names_to = "quantity", values_to = "percent_difference") |>
  mutate(
    source = factor(source, levels = audit$source),
    quantity = recode(
      quantity,
      SSB = "SSB",
      N = "Numbers at age",
      F = "Fishing mortality",
      fishery_selectivity = "Fishery selectivity",
      catch = "Predicted catch"
    )
  ) |>
  ggplot(aes(source, percent_difference, color = quantity, group = quantity)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.2) +
  scale_y_log10(labels = label_number()) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = NULL, y = "Maximum absolute difference (%)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 32, hjust = 1),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

## Objective and gradient change

The initial standalone repository retained the bridged state trajectories but
fixed steepness at `0.67` and changed the parameter map. The NLL discrepancy
increased from approximately `0.0041` to `0.5064`, and the maximum gradient
increased from approximately (6.4\times10^{-6}) to `4.68`.

This is a configuration regression rather than a population-dynamics
regression. Bridge mode must retain every ADMB parameter value and the original
map; development assumptions belong in a separate configuration.

## First state-trajectory regression

Commit `7db7217e4016c2f9d8f6b988fb308ca456b753a8` is the first tested revision
with large state and prediction discrepancies. It added
`cap_old_age_log_selectivity()` and applied it to fishery-selectivity form 0.
Ages 11 and older were set equal to age 11 and the curve was recentered, but the
ADMB reference was not changed.

That single structural change propagated through fishery selectivity, F,
numbers at age, predicted catch, SSB, and the likelihood. Form 0 must remain
unchanged in bridge mode. Old-age tying should be an explicitly named
sensitivity with its own matching ADMB reference if exact equivalence is
required.

# Required development sequence

1. Retain this report and `tests/test_sept_2025_bridge.R` as the deterministic gate.
2. Reintroduce selectable fishery forms without changing form 0; rerun the gate.
3. Add double-logistic and 2D-AR1 sensitivities one at a time; rerun the gate after each.
4. Add old-age tying only as a separate option and keep it disabled in bridge mode.
5. Reintroduce profiling helpers and rerun the deterministic gate.
6. Reintroduce SparseNUTS only after its underlying deterministic model passes.
7. Implement Grant Adams' `m23_rceattle_full` as a separately named RTMB configuration with independent artifacts and bridge thresholds.

# Reproducibility

```{r}
#| label: bridge-test-output
#| echo: true
#| code-fold: false

cat(paste(bridge_output, collapse = "\n"))
```

The bridge can be checked independently of this document with:

```bash
Rscript tests/test_sept_2025_bridge.R
```

Verify the archived source and ADMB artifact checksums with:

```bash
shasum -a 256 -c bridge/sept_2025.sha256
```

The active RTMB profiling work remains isolated on `agent/rtmbprof-stage1`.
This restored report was developed on `bridge/restore-sept-2025` so the dirty
profiling worktree was not modified.
